---
title: "PA1_template REV XX"
author: "DanBeaton"
date: "3/26/2017"
output: 
  html_document: 
    fig_caption: yes
    keep_md: yes
---

#Data Science Course Reproducible Research, Week2.


This assignment analyses the dataset obtained from personal activity monitors.  
Load the dataset into the working directory:

```{r global_options}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figure-html/',
                      warning=FALSE, message=FALSE)
```

```{r, echo=TRUE}

activity = read.csv("activity.csv", header = T)

```

Data were collected at 5 minute intervals spanning the dates 2012-10-01 to 2012-11-30
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.


```{r, echo=TRUE}
head(activity)
tail(activity)

dim(activity)

```


# Plot a Histogram of the Number of Steps Taken per Day



```{r aggregate the data to plot a histogram, echo=TRUE}

dfd = aggregate(steps ~ date, data = activity, sum)
dim(dfd)

```

```{r plot histogram 1, echo=TRUE, include = TRUE}

library(ggplot2)


ggplot(dfd, aes(steps)) +
  geom_histogram(fill = "orchid", alpha = 0.7, color = "grey70", binwidth = 2500) +
  theme(plot.title=element_text(size = 20),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.title.y = element_text(size = 17),
        axis.title.x = element_text(size = 17)) +
  labs(y = "Frequency", x = "Steps per Day") + 
  ggtitle("Figure 1")


```

# Calculate the mean and median number of steps taken per day

For this part of the assignment, you can ignore the missing values in the dataset.
Calculate the total number of steps taken per day


```{r mean, echo=TRUE}

mean.steps = mean(dfd$steps)
mean.steps

```

```{r median, echo=TRUE}

median.steps = median(dfd$steps)
median.steps

```

The mean number of steps per day is `r mean.steps`.
The median number of steps per days is `r median.steps`.

The mean and median are basically the same value.

## What is the average daily activity pattern?

Make a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r interval, echo=TRUE}
library(plyr)
df.ave = ddply(activity, ~interval, summarise, mean = mean(steps, na.rm = T))

```

```{r plot of intervals, echo=TRUE}

ggplot(df.ave, aes(interval, mean)) +
  geom_line(color = "red", lwd = 2) +
  theme(plot.title=element_text(size = 20),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.title.y = element_text(size = 17),
        axis.title.x = element_text(size = 17)) +
  labs(x= "interval", y = "mean of the interval") +
  ggtitle("Figure 2")


```






### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r maximum 5 minute interval, echo=TRUE}

df.max.int = df.ave$interval[which.max(df.ave$mean)]
df.max.int

max.int = max(df.ave$mean)
max.int

```

The interval with the maximum actitivty = `r df.max.int`.
This interval corresponds to an actvity of `r max.int`.


# Imputing Missing Values

###Calcluate the number of missing values "NA"

```{r sum missing values, echo=TRUE}

length.act = length(activity$steps)
numb.na = sum(is.na(activity$steps))
numb.na

```

The total number of missing values in the 'activity' dataset is `r numb.na`.

###Convert the NAs to the mean value of the total activity

```{r convert nas, echo=TRUE}
calcMean = function(x) replace(x, is.na(x), mean(x, na.rm = T))

activity2 = ddply(activity, ~interval, transform, steps = calcMean(steps))

sum(is.na(activity2$steps))

```




#Replot the histogram with the NAs replaced with the mean values


```{r calulate NAs imputing as mean , echo=TRUE}

df.ave2 = ddply(activity2, ~interval, summarise, mean = mean(steps, na.rm = T))
dfd2 = aggregate(steps ~ date, data = activity, sum)

```

```{r histogram2 nas imputed as means, echo=TRUE}

ggplot(dfd2, aes(steps)) +
  geom_histogram(fill = "chartreuse3", alpha = 0.7, color = "grey70", binwidth = 2500) +
  theme(plot.title=element_text(size = 20),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.title.y = element_text(size = 17),
        axis.title.x = element_text(size = 17)) +
  labs(y = "Frequency", x = "Steps per Day") + 
  ggtitle("Figure 3")

```




For this part of the assignment, you can ignore the missing values in the dataset.
Calculate the total number of steps taken per day


```{r mean adjusted, echo=TRUE}

mean.steps2 = mean(dfd2$steps)
mean.steps2

(abs(mean.steps - mean.steps2)/mean.steps)*100

```

```{r median adjusted, echo=TRUE}

median.steps2 = median(dfd2$steps)
median.steps2


(abs(median.steps - median.steps2)/mean.steps)*100
```

The mean number of steps per day with NAs imputed to the mean value is `r mean.steps2`.
The median number of steps per days with NAs imputed to the mean value is `r median.steps2`.

####Do these values differ from the estimates from the first part of the assignment? 

The adjusted mean and median values are basically the same values as the un-adjusted 
mean and median values.


##What is the impact of imputing missing data on the estimates of the total daily number of steps?

Because the mean and median values are similar to each other in the original dataset, 
the imputed NAs replaced with the mean values, do not alter the overall
distribution of the activity levels within each of the 5 minute inervals.


#Are there differences in activity patterns between weekdays and weekends?

```{r weekday vs weekends, echo=TRUE}
#convert the 'date' colum to a date format

str(activity)
activity$date = as.Date(activity$date)
str(activity)

# create labels for the weekedays

library(dplyr)

# create a new colum with the levels 'weekday' and 'weekend'

activity = mutate(activity, day = weekdays(date))



# subset

for (i in 1:length(activity$day)) {
  
  if (activity[i,4] == "Saturday" | activity[i,4] == "Sunday" ) {
    activity[i,4] = "weekend"
    }
  
  else {
    activity[i,4] = "weekday"
  }
}


activity$day = as.factor(activity$day)


head(activity)


```


### Create line plots comparing activity levels for weekends and weekdays


```{r weekend vs weekday, echo=TRUE}

weekday = subset(activity, day == "weekday")
weekend = subset(activity, day == "weekend")

df.wd = ddply(weekday, ~interval, summarise, mean = mean(steps, na.rm = T))
df.we = ddply(weekend, ~interval, summarise, mean = mean(steps, na.rm = T))

```

```{r weekend and weekday plot, echo=TRUE}

g2.line = ggplot() +
  geom_line(data =df.wd, aes(x = interval, y = mean)) + 
  geom_line(data =df.we, aes(x = interval, y = mean), color = "seagreen4") + 
  theme(legend.text = element_text(size = 22)) +
  theme(plot.title=element_text(size = 20),
        axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.title.y = element_text(size = 17),
        axis.title.x = element_text(size = 17)) +
  labs(x= "interval", y = "mean of the interval") +
  annotate("text", x = 500, y = 120, label = "Weekday", size = 8) +
  annotate("text", x = 1500, y = 200, label = "Weekend", size = 8, color = "seagreen4") +
  ggtitle("Figure 4")

g2.line

```

